\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-10 analysis/paper/paper.Rmd
edit litt review

# Literature review
To conduct our review on current methods available in survival analysis, a scraper was coded to examine all packages in the survival task view on CRAN [-@survTaskView]. There was a total of 259 packages, many of which are not relevant to the problems casebase addresses. To filter these packages, a scraper was used to search for competing risks, cumulative incidence functions, non- proportional hazards and penalization. This reduced our set of 259 packages down to 60. From here, we manually examined each package to determine which ones are comparable to casebase, with either a different approach to the same problems or future features we would like to add to casebase.

These packages can belong to three main categories: non-proportional hazards, regularization and competing risks. Besides these three main categories, there are extra features of interest as well. Table \ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those, \pkg{glmnet} [-@regpathcox]???, \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche] introduce penalization in their model fitting. \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package] include penalization, but they implement the extended cox model permitting intuitive non proportional hazard modeling while the previous four do not. In terms of penalization, the benefits of \pkg{casebase} comes from its flexibility. Linear models have a wide tool set, considerably larger than those made for Cox regression. <!--(*reference for number of packages between their task views, regression is spread out across Multivariate, SocialSciences, Robust among others. The focus of this task view is on Optimization Infrastructure Packages , General Purpose Continuous Solvers , Mathematical Programming Solvers , and Specific Applications in Optimization , or Multi Objective Optimization*)--> \pkg{casebase} permits many regularization tools developed for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in a handful of packages. Of interest are \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas \pkg{flexsurv} and \pkg{survival} contain implementations for users to supply any distribution they would like [-@survival-package], [-@flexsurv]. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can use splines, permitting a flexible fit over time [-@flexsurv], [-@smoothHazard], [-@scheike2014estimating], [-@clements_liu]. Though \pkg{casebase} is limited to three distributions at the moment, it also uses splines allowing for distributions that may not be consistent over survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@survival-package]. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot. Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and \pkg{survival} have implementations for cumulative incidence [-@sharabiani_mahani_2019], [-@flexsurv], [-@scheike2014estimating], [-@gerds_blanche], [-@clements_liu], [-@survival-package]. Since \pkg{casebase} is parametric, the baseline hazard is known and requires integration to retrieve the cumulative incidence. This is similar to the approaches of \pkg{CFC} and \pkg{flexsurv} [-@sharabiani_mahani_2019], [-@flexsurv], while the other packages use semi-parametric approaches to retrieve the baseline after fitting [-@scheike2014estimating], [-@gerds_blanche]. \pkg{rstpm2} and \pkg{survival} use both for their respective parametric and semi-parametric model implementations [-@clements_liu], [-@survival-package].

@@

\textbf{casebase} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
$$S(t) = \exp\left(-\int_0^t h(u;X) du\right).$$
The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as

$$ CI(t) = 1 - S(t).$$

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.  Imputation was conducted using the default methods in the mice package in R, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk preditions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the casebase package.

@@

temp          & Temperature (celcius) Day 3              &                 & double           & 0            \\

@@

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
#demonstrating incidence density after imputation
plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
```
Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1.
```{r CBhazard, echo = TRUE, eval = eval_cs3}
#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
#to determine the hazard function
cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                             formula_time = ~d.time,alpha=1,ratio=100)
```
Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
```{r CBabsolute, echo = TRUE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
                              s="lambda.1se",method=c("numerical"))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The regularized cox regression absolute risk curve required two extra steps to retrive. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coeficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

@@

Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occured.

@@

>>>>

# Existing packages

Survival analysis is an important branch of applied statistics and epidemiology. Accordingly, there is a vast ecosystem of \code{R} packages implementing different methods. In this section, we describe how the functionalities of \pkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's task view on survival analysis reveals that there are over 250 packages related to survival analysis [-@survTaskView]. For the purposes of this article, we restricted our description to packages that implement at least one of the following features: parametric modeling, non-proportional hazard models, competing risk analysis, penalized estimation, and cumulative incidence curve estimation. By searching for appropriate keywords in the \code{DESCRIPTION} file of these packages, we found 60 relevant packages. These 60 packages were further reduced through manual curation to a subset of 14 packages. These packages appears in Table \ref{tab:surv-pkgs}, along with some of the functionalities they offer. \textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}

Several packages implement penalized estimation for the Cox model: \pkg{glmnet} [-@regpathcox], \pkg{glmpath} [-@park_hastie], \pkg{penalized} [-@l1penal], \pkg{RiskRegression} [-@gerds_blanche]. Moreover, some packages also include penalized estimation in the context of Cox models with time-varying coefficients: \pkg{CoxRidge} [-@perperoglou], \pkg{rstpm2} [-@clements_liu] and \pkg{survival} [-@survival-package]. On the other hand, \pkg{casebase} provides penalized estimation in the context of parametric hazards. To our knowledge, this is the only package to offer this functionality. \textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@sharabiani_mahani_2019], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival}. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families (see Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear model of covariates (including time). Therefore, our package allows the user to model the effect of time using splines, and through interaction terms involving covariates and time, it also allows user to fit time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv}, \pkg{mets} and \pkg{survival} contain implementations for competing risks. The differentiating factor between these packages and \pkg{casebase} is that \pkg{casebase} handles competing risks through multiple logistic regression. \textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

Finally, several packages include functions to estimate the cumulative incidence function. The corresponding methods generally fall into two categories: transformation of the estimated hazard function, and semi-parametric estimation of the baseline hazard. The first category broadly corresponds to parametric survival models, where the full hazard is explictly modeled. Using this estimate, the survival function and the cumulative incidence function can be obtained using their functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below). Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows this approach for both single-event and competing-event analyses. The second category outlined above broadly corresponds to Cox models. These models do not model the full hazard function, and therefore the baseline hazard needs to be estimated separately in order to estimate the survival function. This is achieved using semi-parametric estimators (e.g. Breslow's estimator). Packages that implement this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As mentioned in the introduction, a key distinguishing factor between these two approaches is that the first category leads to smooth estimates of the cumulative incidence function, whereas the second category produces estimates in the form of step-wise functions. This was one of the main motivations for introducting case-base sampling in survival analysis.

@@

\textbf{casebasee} \cite{hanley2009fitting}       & x                        & x                         & x                     & x                & x                   &                          &                                  & x                                \\

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all the familiar summary and diagnostic functions are available: \code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our package provides one more functionality: it computes risk functions from the model fit. For the case of a single event, it uses the familiar identity
\begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t h(u;X) du\right).
\end{equation}
The integral is computed using either the \code{stats::integrate} function or Monte-Carlo integration. The risk function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT dataset tracks death for individuals who are considered seriously ill within a hospital.  Imputation was conducted using the default methods in the mice package in R, available on CRAN. Before imputation, there was a total of 1000 individuals and 35 variables. After imputation, we have 474 observations where 67.5% of individuals in the study experience the event of interest. The SUPPORT dataset will be used to demonstrate regularization within casebase (REF), while comparing their absolute risk predictions for a new individual. A description of each variable can be found in Table~\ref{tab:support1} and a breakdown of each categorical variable in Table~\ref{tab:support2}. The data was downloaded from the main author's website, and it is also available as part of the casebase package.

@@

temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\

@@

```{r poptimePostImputation, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:ptPostI} Population-time plot before imputation. Red dots demonstrate death, while each gray dot makes up all the person moments spent in the study."}
#demonstrating incidence density after imputation
plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
```
Our main objective is to compute the absolute risk of death for a given set of covariates, using regularization when fitting our model. First, we fit a smooth hazard to the data; for the sake of this example, we opted for a linear term for time. As casebase makes use of the glmnet package, we interact with the fitsmoothhazard.fit function using a matrix interface, where y contains the time and event variables and x contains all other variables we would like to include in the model. For the SUPPORT dataset, all factor variables were converted into dummy variables. Here, we used lasso penalization by setting alpha to 1.
```{r CBhazard, echo = TRUE, eval = eval_cs3}
#Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
#to determine the hazard function
cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
                             formula_time = ~d.time,alpha=1,ratio=100)
```
Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.
```{r CBabsolute, echo = TRUE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
                              s="lambda.1se",method=c("numerical"))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The regularized cox regression absolute risk curve required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires that a survival object skeleton of the selected variables, so that the survival package tools can be used to retrieve the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

@@

Recall the setting of this study: patients were admitted to the Stanford program after meeting with their physician and determining that they were unlikely to respond to other forms of treatment. After enrollment, the program searches for a suitable donor for the patient, which can take anywhere between a few days to almost a year. We are interested in the effect of a heart transplant on survival; therefore, the patient is considered exposed only after the transplant has occurred.

@@

\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval censoring, whereas \pkg{casebase} currently cannot. Interval censoring is a simple feature to implement and is a potential improvement to add to \pkg{casebase}.

### 2019-12-10 analysis/paper/paper.tex
edit litt review

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

To conduct our review on current methods available in survival analysis,
a scraper was coded to examine all packages in the survival task view on
CRAN \citeyearpar{survTaskView}. There was a total of 259 packages, many
of which are not relevant to the problems casebase addresses. To filter
these packages, a scraper was used to search for competing risks,
cumulative incidence functions, non- proportional hazards and
penalization. This reduced our set of 259 packages down to 60. From
here, we manually examined each package to determine which ones are
comparable to casebase, with either a different approach to the same
problems or future features we would like to add to casebase.

These packages can belong to three main categories: non-proportional
hazards, regularization and competing risks. Besides these three main
categories, there are extra features of interest as well. Table
\ref{tab:surv-pkgs} summarizes the findings.

Many packages exist that implement semi-parametric Cox models. Of those,
\pkg{glmnet} \citeyearpar{regpathcox}???, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche} introduce penalization
in their model fitting. \pkg{CoxRidge} \citeyearpar{perperoglou},
\pkg{rstpm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package} include penalization, but they implement
the extended cox model permitting intuitive non proportional hazard
modeling while the previous four do not. In terms of penalization, the
benefits of \pkg{casebase} comes from its flexibility. Linear models
have a wide tool set, considerably larger than those made for Cox
regression. \pkg{casebase} permits many regularization tools developed
for linear regression to be repurposed within a survival context.

Parametric models permit non-proportional hazards and are implemented in
a handful of packages. Of interest are \pkg{CFC}
\citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu} and \pkg{survival}
\citeyearpar{survival-package}. \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas \pkg{flexsurv} and
\pkg{survival} contain implementations for users to supply any
distribution they would like \citeyearpar{survival-package},
\citeyearpar{flexsurv}. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets}
and \pkg{rstpm2} can use splines, permitting a flexible fit over time
\citeyearpar{flexsurv}, \citeyearpar{smoothHazard},
\citeyearpar{scheike2014estimating}, \citeyearpar{clements_liu}. Though
\pkg{casebase} is limited to three distributions at the moment, it also
uses splines allowing for distributions that may not be consistent over
survival time.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks \citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{survival-package}. The
differentiating factor between these packages and \pkg{casebase} is that
\pkg{casebase} handles competing risks through multiple logistic
regression. \pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can
handle interval censoring, whereas \pkg{casebase} currently cannot.
Interval censoring is a simple feature to implement and is a potential
improvement to add to \pkg{casebase}.

Of all the packages listed in Table \ref{tab:surv-pkgs}, \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, \pkg{RiskRegression}, \pkg{rstpm2} and
\pkg{survival} have implementations for cumulative incidence
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv},
\citeyearpar{scheike2014estimating}, \citeyearpar{gerds_blanche},
\citeyearpar{clements_liu}, \citeyearpar{survival-package}. Since
\pkg{casebase} is parametric, the baseline hazard is known and requires
integration to retrieve the cumulative incidence. This is similar to the
approaches of \pkg{CFC} and \pkg{flexsurv}
\citeyearpar{sharabiani_mahani_2019}, \citeyearpar{flexsurv}, while the
other packages use semi-parametric approaches to retrieve the baseline
after fitting \citeyearpar{scheike2014estimating},
\citeyearpar{gerds_blanche}. \pkg{rstpm2} and \pkg{survival} use both
for their respective parametric and semi-parametric model
implementations \citeyearpar{clements_liu},
\citeyearpar{survival-package}.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \[S(t) = \exp\left(-\int_0^t h(u;X) du\right).\] The integral
is computed using either the \code{stats::integrate} function or
Monte-Carlo integration. The risk function (or cumulative incidence
function) is then defined as

\[ CI(t) = 1 - S(t).\]

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk preditions for a new individual. A description of each variable can
be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown of
each categorical variable in Table\textasciitilde{}\ref{tab:support2}.
The data was downloaded from the main author's website, and it is also
available as part of the casebase package.

@@

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The regularized cox regression
absolute risk curve required two extra steps to retrive. Coxnet from the
glmnet package has tools to fit a regularized cox model, but requires
that a survival object skeleton of the selected variables, so that the
survival package tools can be used to retrive the absolute risk. We
handle this by fitting a regularized cox model with the glmnet package,
and a cox model from the survival package with the selected variables
from the regularized fit. This second model serves as a skeleton that
permits the use of survival package functions. The coefficients from the
regularized model will replace the coeficients in our skeleton. the
resulting model will have the correct coefficients when integrating, but
will have an incorrect standard error. For the purposes of this case
study, we are only interested in the absolute risk curve itself and not
the standard error.

@@

This report was generated on 2019-12-08 12:31:41 using the following
computational environment and dependencies:

@@

#>  date     2019-12-08                  

@@

\begin{CodeOutput}
#> Local:    review-max /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [69663bb] 2019-12-08: Merge branch 'review-max' of https://github.com/sahirbhatnagar/cbpaper into review-max
\end{CodeOutput}
\end{CodeChunk}

>>>>

\hypertarget{existing-packages}{%
\section{Existing packages}\label{existing-packages}}

Survival analysis is an important branch of applied statistics and
epidemiology. Accordingly, there is a vast ecosystem of \code{R}
packages implementing different methods. In this section, we describe
how the functionalities of \pkg{casebase} compare to these packages.

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were further reduced
through manual curation to a subset of 14 packages. These packages
appears in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.
\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{sharabiani_mahani_2019}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}. The types of
models they allow vary for each package. For example, \pkg{SmoothHazard}
is limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of
time using splines, which allows flexible modeling of the hazard
function. Moreover, \pkg{flexsurv} has the ability to estimate both
scale and shape parameters for a variety of parametric families (see
Table \ref{tab:dists}). As discussed above, \pkg{casebase} can model any
parametric family whose log-hazard can be expressed as a linear model of
covariates (including time). Therefore, our package allows the user to
model the effect of time using splines, and through interaction terms
involving covariates and time, it also allows user to fit time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Of the methods mentioned so far, only \pkg{CFC}, \pkg{flexsurv},
\pkg{mets} and \pkg{survival} contain implementations for competing
risks. The differentiating factor between these packages and
\pkg{casebase} is that \pkg{casebase} handles competing risks through
multiple logistic regression.
\textcolor{red}{JESSE: There's a literature review of packages for competing-risk analysis in the paper for \pkg{CFC}. Could you incorporate some of their discussion, with proper citations? The biggest omission right now is \pkg{cmprsk}, which we actually use below.}

Finally, several packages include functions to estimate the cumulative
incidence function. The corresponding methods generally fall into two
categories: transformation of the estimated hazard function, and
semi-parametric estimation of the baseline hazard. The first category
broadly corresponds to parametric survival models, where the full hazard
is explictly modeled. Using this estimate, the survival function and the
cumulative incidence function can be obtained using their functional
relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI} below).
Packages including this functionality include \pkg{CFC}, \pkg{Flexsurv},
\pkg{mets}, and \pkg{survival}. Our package \pkg{casebase} also follows
this approach for both single-event and competing-event analyses. The
second category outlined above broadly corresponds to Cox models. These
models do not model the full hazard function, and therefore the baseline
hazard needs to be estimated separately in order to estimate the
survival function. This is achieved using semi-parametric estimators
(e.g.~Breslow's estimator). Packages that implement this approach
include \pkg{RiskRegression}, \pkg{rstpm2}, and \pkg{survival}. As
mentioned in the introduction, a key distinguishing factor between these
two approaches is that the first category leads to smooth estimates of
the cumulative incidence function, whereas the second category produces
estimates in the form of step-wise functions. This was one of the main
motivations for introducting case-base sampling in survival analysis.

@@

Once a model-fit object has been returned by \code{fitSmoothHazard}, all
the familiar summary and diagnostic functions are available:
\code{print}, \code{summary}, \code{predict}, \code{plot}, etc. Our
package provides one more functionality: it computes risk functions from
the model fit. For the case of a single event, it uses the familiar
identity \begin{equation}\label{eqn:surv}
S(t) = \exp\left(-\int_0^t h(u;X) du\right).
\end{equation} The integral is computed using either the
\code{stats::integrate} function or Monte-Carlo integration. The risk
function (or cumulative incidence function) is then defined as
\begin{equation}\label{eqn:CI}
CI(t) = 1 - S(t).
\end{equation}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk predictions for a new individual. A description of each variable
can be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown
of each categorical variable in
Table\textasciitilde{}\ref{tab:support2}. The data was downloaded from
the main author's website, and it is also available as part of the
casebase package.

@@

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The regularized cox regression
absolute risk curve required two extra steps to retrieve. Coxnet from
the glmnet package has tools to fit a regularized cox model, but
requires that a survival object skeleton of the selected variables, so
that the survival package tools can be used to retrieve the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coefficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

@@

This report was generated on 2019-12-10 12:28:40 using the following
computational environment and dependencies:

@@

#>  date     2019-12-10                  

@@

\begin{CodeOutput}
#> Local:    review /home/mturgeon/Documents/git_repositories/cbpaper
#> Head:     [4435e71] 2019-12-08: Merge branch 'review-max'
\end{CodeOutput}
\end{CodeChunk}
